A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-12-06 Simulation/Writeup/NoteConvergence.Rnw
Update pdf to review
Feedback Gerko on Introduction.

At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}. Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which can yield wildly invalid results \citep{buur18}. A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}. MI is an iterative algorithmic procedure in which each missing data point is `guessed' (i.e. imputed) several times. The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness. Therefore, MI can provide valid inferences despite missing information.

\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be correct. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it requires more observations than strictly required. Since both of these situations are undesirable, imputations and their variability should be properly evaluated. Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \proglang{R} package \pkg{mice} \citep{mice}. 
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The aim of this note is to provide guidelines for the evaluation of imputations to applied researchers and statisticians who use MI methods. More specifically, this note addresses the question: `How to diagnose convergence of the MI algorithm implemented in \pkg{mice}?'.

The convergence properties of \pkg{mice} are investigated by means of model-based simulation in \proglang{R} \citep{R}. All programming code used in this note is available on Github repository \href{https://github.com/gerkovink/shinyMice}{`ShinyMICE'}. The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms. These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. This note therefore follows notation and conventions of \cite{mice}. Deviations from the 'original' notation by \cite{rubin87}, are described in \citep[\S~XYZ]{buur18}.

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. This paper only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.

In this note, the terms `unobserved' and `missing' data are used interchangeably. Both refer to $Y_{mis}$, e.g. `$NA$' values in \proglang{R} or `999' in \proglang{SPSS}. The terms 'incomplete' or 'observed' data denote $Y_{obs}$. Incomplete data is the starting point of the multiple imputation procedure. Figure \ref{fig:steps} provides a schematic overview of the steps involved with MI.

\begin{figure}
\label{fig:steps}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=left:Incomplete data] { };
\node[main] (mids) [right=of data] { };
\node[main] (mids2) [above=of mids] { };
\node[main] (mids3) [below=of mids,label=below:Imputed data] { };
\node[main] (mira) [right=of mids] {};
\node[main] (mira2) [above=of mira] { };
\node[main] (mira3) [below=of mira,label=below:Analysis results] { };
\node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
\caption{Scheme of main steps in multiple imputation (Adapted from \cite[\S~1.4.1]{buur18})}
\end{figure}

@@

This note is primarily concerned with the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain` of potential values is considered. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate value that a chain lands on is imputed, and hence used in the next steps (analysis and pooling). The collection of values (or summary statistics) across iterations of one of the $m$ imputations will be referred to as `imputation chain'.

\textbf{ADD THIS? The process of deliberately creating missingness in an otherwise complete data set is referred to as `amputing' the data.}

\textbf{[explain difference MI, FCS, mice]}

\textbf{[Explain `simulation condition' (the number of iterations the algorithm has before imputing), ...?]}

% \begin{figure}[h]
%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
%         \includegraphics{steps.png}
%   }
%   \caption{Scheme of main steps in multiple imputation (Adapted from \cite[par.~1.4.1]{buur18})}
%     \label{resultssim}
% \end{figure}

% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.

% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample. We define the response indicator R as an n x p 0-1 matrix. The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ... , n and j = 1 , ... , p . If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ... The observed data are collectively denoted by Y o b s . The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 . When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing. The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors. Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself. \cite[par.~2.2]{buur18}

@@

%% -- Theoretical Background ---------------------------------------------------------------
\subsection{Theoretical Background (NOG NIET AF VANAF HIER)} \label{sec:background}

@@

We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}. % In general: "In general, you cannot know for sure if your chain has converged. But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
Convergence has two interpretations \textbf{[look up source!!!]}: mixing between chains and stability over iterations within chains. Ideally, we want the chains to intermingle nicely, without trends within chains.  Diagnose non-convergence in the mixing sense: $\widehat{R}$. Diagnose non-convergence in the stability sense: auto-correlation.

@@

In the latter interpretation of convergence, convergence is interpreted as stability over iterations, or non-recurring. Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. Auto-correlation above the confidence level \textbf{[explain that this is a critical value for correlation based on alpha .05]} indicates dependence within chains. Also, possible to use MC error, or Geweke statistic. But that is outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$.

The potential scale reduction factor $\widehat{R}$ tells us how much variance of an estimate could be shrunken down by running the chains infinitely long \cite{gelm92}. That then tells us something about how dependent the chains are on the starting values. If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word). $\widehat{R}$ is then equal to one. The conventional threshold was $\widehat{R} < 1.2$ is acceptable. Most recent guideline by \cite{veht19} is more stringent: $\widehat{R} < 1.01$.

We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest. $\widehat{R}$ is computed as follows: % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}

@@

@@

A fundamental assumption of MICE is convergence of the algorithm. MICE is a sort of Gibbs sampler, a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.

Why is $\widehat{R}$ not an appropriate diagnostic to evaluate MI data? $\widehat{R}$ is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the $m$ imputation chains are 'far away' from the distribution that the chains are converging to. With MI procedures, initial values are chosen randomly from the observed data. This means that one initial value is necessary for each imputation chain.

@@

Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \citep{broo98}. %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. 
This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms. An empirical finding, however, shows that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations. %that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50. Therefore, the aim is to replicate empirical finding Lacerda et al.

@@

In this study, 1000 simulations are performed for each of the 100 simulation conditions. The simulation set-up consists of several steps, summarized in the pseudo-code below. The complete \proglang{R} script of the simulation study is available on Github.

@@

\subsection{Data}
The simulated dataset is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression complete data problem. The quantity of scientific interest is the estimated regression coefficient of predictor $X$ on outcome variable $Y$. The linear regression model is $Y \sim X+Z_1+Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates. The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$. Outcome variable $Y$ is deduced from the predictor variables as follows:

\begin{align*}
\begin{pmatrix}X\\
Z_{1}\\
Z_{2}\\
\epsilon
\end{pmatrix} &\sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,&
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\\[2\jot]
Y &=  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon
\end{align*}

@@

$\widehat{R}$ is computed within imputation chains: for each variable, for each simulation condition, for each simulation. The maximum value across variables within the same simulation is reported. Maximum $\widehat{R}$ values per simulation condition are aggregated across simulations.

Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration. For now, we only consider the chain means as scalar of interest $\theta$. The AC of the variable with the highest absolute AC value is reported per simulation. These values are then averaged per simulation condition.

@@

\subsection{Simulation diagnostics}
The simulation diagnostics are as recommended by \cite{buur18}. Namely, average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations. %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}. For now, this is outside of the scope of this study. % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

Figures \ref{fig:conv} and \ref{fig:sim} shows the results over 1000 simulations. A subset of simulation conditions is presented in Table \ref{tab:results}.

@@

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{Figures/Rplot.pdf}
  }
  \caption{Simulation diagnostics over 1000 MCMC simulations.}
    \label{fig:sim}
\end{figure}

Evaluate simulation diagnostics one by one: bias, empirical SE, confidence interval width, and coverage rate. Then evaluate convergence diagnostics $\widehat{R}$ and auto-correlation.

@@

\textbf{$\widehat{R}$}. Steep drop down up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations (see online appendix for the full table of results).

@@

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending up-to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.

\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym43}, namely \Sexpr{round(mean(dat[, "Cov. rate"]), 3)}.

>>>>

At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}. Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results \citep{buur18}. A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by \cite{rubin87}. MI is an iterative algorithmic procedure in which each missing data point is `imputed' (i.e. filled in) several times. The variability between imputations is used to reflect how much uncertainty in the inference is introduced by the missingness. Therefore, MI can provide valid inferences despite missing information.

\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be properly represented. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power. Since both of these situations are undesirable, imputations and their variability should be properly evaluated. Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}. 
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The aim of this research project is to investigate the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). This note addresses the question: `How to diagnose convergence of the multiple imputation algorithm implemented in \pkg{mice}?'.

The convergence properties of the MICE algorithm are investigated by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI methods. These guidelines will be implemented in the interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. This note therefore follows notation and conventions of \cite{mice}. Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~XYZ]{buur18}.

Let $Y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The collection of observed data values in $Y$ is denoted as $Y_{obs}$; the missing part of $Y$ is referred to as $Y_{mis}$. Response indicator $R$ shows whether a data value in $Y$ is missing or observed. The relation between $R$, $Y_{obs}$, and $Y_{mis}$ determines the missingness mechanism. This note only considers a `missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $Y$.

In this note, the terms `unobserved' and `missing' data are used interchangeably. Both refer to $Y_{mis}$, e.g. `\texttt{NA}' values in \proglang{R} or `\texttt{999}' in \proglang{SPSS}. The terms `incomplete' or `observed' data denote $Y_{obs}$. Incomplete data is the starting point of the multiple imputation procedure. Figure \ref{fig:steps} provides a schematic overview of the steps involved with MI.

\begin{figure}
\label{fig:steps}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=left:Incomplete data] { };
\node[main] (mids) [right=of data] { };
\node[main] (mids2) [above=10mm of mids] { };
\node[main] (mids3) [below=10mm of mids,label=below:Imputed data] { };
\node[main] (mira) [right=of mids] {};
\node[main] (mira2) [above=10mm of mira] { };
\node[main] (mira3) [below=10mm of mira,label=below:Analysis results] { };
\node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
\caption{Scheme of the main steps in multiple imputation ($m = 3$). Adapted from \citep[\S~1.4.1]{buur18}.}
\end{figure}

@@

This note focusses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is considered. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate value that a chain lands on is imputed, and hence used in the next steps (analysis and pooling). The collection of values (or summary statistics) across iterations of each of the $m$ imputations will be referred to as an `imputation chain'.

% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.
% "Let Y denote the n x p matrix containing the data values on p variables for all n units in the sample. We define the response indicator R as an n x p 0-1 matrix. The elements of Y and R are denoted by y i j and r i j , respectively, where i = 1 , ... , n and j = 1 , ... , p . If y i j is observed, then r i j = 1 , and if y i j is missing, then r i j = 0 ." ... The observed data are collectively denoted by Y o b s . The missing data are collectively denoted as Y m i s , and contain all elements y i j where r i j = 0 . When taken together Y = ( Y o b s , Y m i s ) contain the hypothetically complete data." \cite[par.~2.2]{buur18}
% The data are said to be MCAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | $\psi$ ) (2.1) so the probability of being missing depends only on some parameters $\psi$ , the overall probability of being missing. The data are said to be MAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) = Pr ( R = 0 | Y o b s , $\psi$ ) (2.2) so the missingness probability may depend on observed information, including any design factors. Finally, the data are MNAR if Pr ( R = 0 | Y o b s , Y m i s , $\psi$ ) (2.3) does not simplify, so here the probability to be missing also depends on unobserved information, including Y m i s itself. \cite[par.~2.2]{buur18}

@@

%% -- Theoretical Background ---------------------------------------------------------------
\subsection{Theoretical Background} \label{sec:background}

The method of multiple imputation The properties of multiple imputation procedures have been studied

@@

We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}. % In general: "In general, you cannot know for sure if your chain has converged. But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
Convergence has two components \textbf{[look up source!!!]}: mixing between chains and stability over iterations within chains. Ideally, both components will be satisfied: there are no trends within chains, and the chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Diagnose non-convergence in the mixing sense: $\widehat{R}$. Diagnose non-convergence in the stability sense: auto-correlation (AC). Also, it is possible to compute MC errors, or the Geweke statistic \citep{XYZ}. All of these methods evaluate the convergence of univariate scalars of interest (e.g., chain mean or chain variance). The convergence of multivariable statistics (i.e. relations between such scalars) may be evaluated via the eigenvector decomposition method proposed by McKay \citep{XYZ}. These three methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$.

@@

The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down, if the chains were to be infinitely long \cite{gelm92}. That then informs us about how dependent the chains are on the starting values. If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word). $\widehat{R}$ is then equal to one. The conventional acceptable threshold was $\widehat{R} < 1.2$. More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.

We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest. $\widehat{R}$ is defined by \cite{veth19} as follows: % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}

@@

High $\widehat{R}$ would indicate .... low $\widehat{R}$ values indicate ....

The second component of convergence ... interpreted as stability over iterations, or non-recurring. Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. A high auto-correlation indicates dependence within chains. The magnitude of the AC can be interpreted qualitatively, or compared to the critical value of a \textbf{XYZ} distribution with \textbf{XYZ} degrees of freedom.

@@

A fundamental assumption of MICE is convergence of the algorithm. MICE is a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.

$\widehat{R}$ may not be an appropriate diagnostic to evaluate MI data because it assumes over-dispersed initial values. This means that the initial values of the $m$ imputation chains are `far away' from the distribution that the chains are converging to. In the \pkg{mice} algorithm, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. %The initial parameter value is noodzakelijk. Daarom is het lastig om in MI that starten vanuit overdisp state.

@@

Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}. %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. 
This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms. An empirical finding, however, shows that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations. %that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50. Therefore, the aim is to replicate empirical finding Lacerda et al.

@@

To investigate convergence properties of the \pkg{mice} algorithm, different lenghts of imputation chains are considered. this simulation study considers 100 simulation conditions. Each condition has a different maximum number of iterations, varying between one and one-hundred iterations. In this study, 1000 simulation runs are performed for each of the 100 simulation conditions. The simulation set-up consists of several steps, summarized in the pseudo-code below. The complete \proglang{R} script of the simulation study is available on Github.

@@

\subsection{Data}
The simulated dataset is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression complete data problem. The quantity of scientific interest is the estimated regression coefficient of predictor $X$ on outcome variable $Y$. The linear regression model is $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates. The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$. Outcome variable $Y$ is deduced from the predictor variables as $Y =  2 \times X + 0.5 \times Z_1 - 1 \times Z_2 + \epsilon$.

\begin{align*}
\begin{pmatrix}X\\
Z_{1}\\
Z_{2}\\
\epsilon
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\\[2\jot]
\end{align*}

@@

$\widehat{R}$ is computed within imputation chains: for each variable, each simulation condition, and each simulation. The maximum value across variables within the same simulation is reported. Maximum $\widehat{R}$ values per simulation condition are aggregated across simulations.

Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances). The AC of the variable with the highest absolute AC value is reported per simulation. These values are then averaged per simulation condition.

@@

\subsection{Simulation diagnostics}
The simulation diagnostics are as recommended by \cite{buur18}. These comprise of average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations. %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd}. For now, this is outside of the scope of this study. % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

Figures \ref{fig:conv} and \ref{fig:sim} display results of the simulation study aggregated over 1000 repetitions. A subset of simulation conditions is presented in Table \ref{tab:results}. This section describes convergence diagnostics ($\widehat{R}$ and auto-correlation) and simulation diagnostics (bias, confidence interval width and coverage rate) separately.

@@

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{Figures/simulation_diag.pdf}
  }
  \caption{Simulation diagnostics over 1000 MCMC simulations.}
    \label{fig:sim}
\end{figure}

@@

\textbf{$\widehat{R}$}. Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results).

@@

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending up-to thirthy to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.

\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov. rate"]), 3)}\%.

\textbf{add loess line to plots to see when it becomes a flat line.}
